Adding Sensitivity to 21\,cm Inteferometric Probes of Reionization by Optimizing Choice of Baselines

### 2017-05-02 Rotation paper/draft_working.tex
save

%\documentclass[preprint2,numberedappendix,tighten,twocolappendix]{aastex6}  % USE THIS TO MAKE BIB, THEN FORMAT USING EMULATEAPJ
\documentclass[twocolumn,apj,numberedappendix]{emulateapj}
\shorttitle{Adding Sensitivity to 21cm Inteferometric Probes of Reionization}
\shortauthors{Zhang\&Parsons}
\usepackage{float}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[figuresright]{rotating}
%\usepackage{rotating}
\usepackage{natbib}
%\usepackage{pdflscape}
%\usepackage{lscape}
\usepackage{ctable}
\citestyle{aa}
\renewcommand\[{\begin{equation}}
\renewcommand\]{\end{equation}}
\graphicspath{{../figures/}}

@@

\title{Adding Sensitivity to 21cm Inteferometric Probes of Reionization by Optimizing Choice of Baselines}

@@

\begin{abstract}
The observational effort to measure the primordial 21cm power spectrum with radio interferometers require high sensitivity, or signal to noise ratio. We introduce a straightforward and accurate method to extract additional sensitivity from data of redundant radio arrays. Our method efficiently finds optimal baselines to cross correlate, including baselines that are slightly different in length and orientation, and quantifies the sensitivity contributions of given baselines. 
Using the configurations and beams of the latest and final 128-element version of Precision Array for Probing the Epoch of Reionization (PAPER-128), as well as the planned 37, 128, 240 and 350 element versions of the Hydrogen Epoch of Reionization Array (HERA) now under development, we illustrate how our method applies to different arrays and predict the sensitivity improvements of including each baseline pairs. We show that inclusion of non-perfectly redundant baselines 
would increase the sensitivity of PAPER-128 and different configurations of HERA by about to $20\%$ to $60\%$, with HERA-350 benefitting the most from the technique. 
\end{abstract}

@@

The epoch of reionization, or cosmic dawn, represents the last key
stage of our universe's early evolution. Study of this event stands at
the intersection of cosmology and astrophysics. Understanding this
event is important not only because it is a scientific goal
of its own, but also because it can provide crucial information
to fundamental physics of inflation, neutrino mass and phenomenology
of the first stars and galaxies (e.g. \citealt{LiuOpticalDepth, Liu2016b, Mao2008, DEw21cm, Bull2015, Oyama20131186} etc.).

Arguably the most promising observational probe of the epoch of reionization
comes from measurement of the ``spin-flip'' transition of neutral
hydrogen of characteristic wavelength 21cm \citep{Furlanetto2006181,PritchardLoeb}.
While probes such as quasar spectra rely on emission and scattering
of free electrons, and are thus limited to the lower redshifts towards the end of reionization, the 21cm line directly probes the
abundance and distribution of neutral hydrogen, and thus would potentially
shed light on all stages of reionization. Radio interferometric efforts
to measure the 21cm power spectrum has been a top priority in recent years of astronomy.
Current generation instruments include the Precision Array for Probing
the Epoch of Reionization (PAPER) \citep{Ali2015,paper32}, Murchison
Widefield Array (MWA) \citep{Bowman2013, Tingay2013}, Low Frequency Array (LOFAR) \citep{LOFAR}. Next generation instruments such as the Hydrogen Epoch of Reionization
Array (HERA) (e.g. \citealt{HERA,HERAconfiguration,HERABEAM1,HERADISH2}) currently under construction
and the Square Kilometer Array Low (SKA-low) (e.g. \citealt{SKA1}) in planning.

One of the main challenges of observations of cosmic
21cm line, interferometric or global, is foreground contamination. Both sources within our
galaxy, and to a lesser extent distant sources outside of our galaxy emit
radio contamination (via for example synchrotron processes) up to five orders of magnitude stronger than the
reionization signal. There are two common methods to deal with the
foreground contamination in power spectrum measurements. In the first, sometimes called the foreground
removal technique, individual sources are identified and removed in the image domain.
The other technique, commonly referred to as ``foreground avoidance'',
makes use of the fact that most common foreground contaminants have smooth
spectra, and thus is constrained to a Fourier domain ``foreground
wedge'' \citep{wedge1,wedge2}. Thus, contamination can be avoided
by restricting our observation to ``outside the wedge'', where theoretically foreground contamination is limited and noise becomes dominant. The main
challenge of using the avoidance technique is thus the sensitivity, or
the signal to noise ratio. This is the motivation for the design of
the maximum redundancy arrays such as PAPER and HERA. Traditional
imaging arrays are designed to image localized bright sources and thus favor non-redundant patterns for $uv$ coverage. The redundant arrays, on the other hand,
focus on high-sensitivity measurements of the same Fourier mode
with many tightly packed antennas at equal spacing. Since baselines
of the same length and orientation measure the same Fourier modes
on the sky, a maximum redundancy array is able to increase the
signal to noise ratio by averaging over measurements of the same baseline.
\cite{Ali2015} provides the newest upper limit to the power spectrum
measurements with the 64-element version of PAPER (henceforth referred to as PAPER-64). From some models we roughly expect
that the sensitivity required for detection is a factor of 10 away.

This paper explores a new technique to add sensitivity to the power spectrum analysis
of redundant arrays, by making use of the earth's rotation. 
The earth's rotation causes the baselines to pick up different modes of the sky with time. 
This effect is used extensively to improve $uv$ coverage in imaging with minimum redundancy arrays. In
a maximally redundant array, we can make use of the fact that baselines that are slightly different
rotate into each other at a time delay to extract power-spectrum from their respective visibilities. Here we present a method to
improve the sensitivity of the maximum redundancy arrays by including
the $uv$ redundancy of different baselines at a time lag. Recently, \cite{wterm} proposed an algorithmic approach to extract information from near-equivalent baselines, by carefully treating the $w$-term in the fringe pattern to obtain weights for cross-correlation. The algorithm makes use of a time dependent coordinate system ('moving grid') and is thus computationally expensive. We here present a formalism that allows us to compute such weights by simply performing a numerical integral. This allows us to identify good baselines to cross-correlate and predict the weight of sensitivity with speed and accuracy, thereby making the power spectrum estimation from a large number of baseline pairs computationally tractable.

Baselines of the same length and orientation are traditionally called
``redundant baselines'', because they measure the same Fourier mode
in the sky. In order to eliminate confusion and ambiguity, we shall
introduce slightly different terminology. We shall call baselines that
are the same length and orientation ``equivalent baselines'', inspired
by the mathematical notion of equivalency classes. Two equivalent
baselines will be redundant with each other simultaneously at all
times. Non-equivalent baselines can also be partially redundant if their
respected time series is shifted with respect to one another by some delay. In other words, one baseline can be ``rotated into''
another. We call these ``near-equivalent baselines''. Exactly how ``near" is ``near-enough" depends on the array configuration. All non-equivalent baselines contribute non-zero sensitivity to the power spectrum analysis It is our
goal to a) identify the near-equivalent baselines that give good
redundancy, b) to find the optimal time offset\footnote{By optimal we mean the time offset that corresponds to maximal redundancy. } for a given pair of baselines, and
c) to quantify the sensitivity improvement associated with cross multiplying
such a pair of near-equivalent baselines.

The rest of this paper is organized as follows. In section
2 we explain the theoretical basis for this cross multiplication.
In section 3 we present numerical tests of
this technique as well as the expected sensitivity improvement
with this method for HERA and PAPER-128 and with section 4 we conclude.

@@

\caption{The PAPER-128 layout. Each blue dot corresponds to the location of
an antenna. Top panel shows the antenna positions drawn to scale,
bottom panel show the antenna labels and distances, excluding the outrigger
antennas.
The numbering of the antennas the bottom panel are original labels
during instrument assembly and does not bear significant
meaning. Baselines corresponding to the same separations are called
equivalent. In the bottom panel, the two baselines (5,77) and (40,85) are an example of equivalent pair, with separation denoted by sep1,0, for the
antennas are separated by 1 unit east and 0 unit north. Similarly,
the baselines (5,69) and (5,85) are examples
of sep1,1 and sep1,-1, respectively. Note sep1,0 and sep-1,0 for example
are the same baselines and should not be counted twice.}
\label{fig:AntPos}
\end{figure}

We shall use the 128-element PAPER array to demonstrate our method, and extend our results to several HERA configurations in Sec \ref{sec:arrconf}. 
The PAPER array is located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern with antenna labels are show
in Fig. \ref{fig:AntPos}. We see the antenna spacing in North-South
directions are comparatively close (4m), so that baselines such as
0\_44 and 0\_7 are very close to equivalent. In the bottom panel, the two baselines (5,77) and (40,85) are an example of equivalent pair, with separation denoted by sep1,0, for the
antennas are separated by 1 unit east and 0 unit north. Similarly,
the baselines (5,69) and (5,85) are examples
of sep1,1 and sep1,-1, respectively.
Note sep1,0 and sep-1,0 for example are the same baselines and should
not be counted twice. Antennas in purely north-south baselines
are close enough to induce cross-talk, and hence are not suitable
for use. The original PAPER-64 analysis \citep{Ali2015} used three classes of baselines, here
equivalent to 
sep2,0, sep2,1 and sep2,-1 \cite{Ali2015} of PAPER-128. There each of these classes
of baselines are cross multiplied to itself. We shall see that in addition these
baseline classes can be cross multiplied with a time offset.

Given a point source on the sky, each baseline maps the
source to a point in $uv$ plane. As the earth rotates with respect to
the source, the point traces out tracks in the $uv$ plane. 
We show in Fig. \ref{fig:Tracks} $uv$ tracks of PAPER-128 over 4.8 hours, at 0.15GHz, for a source that passes through zenith.

\begin{figure}[H]
\includegraphics[width=\linewidth]{tracks128}
\caption{Tracks of PAPER-128 (grid only, excluding outriggers) for requency $\nu=0.15\text{GHz}$ and a hypothetical source that passes through zenith.
These tracks are traced out over 0.2 sidereal days, or roughly 4.8
hours. Color represents different baselines. 
As the earth rotates, tracks are traced out counterclockwise. }
\label{fig:Tracks}
\end{figure}

@@

We take the visibility as commonly defined in the literature (e.g.
\citealt{first-paper}): 
\begin{equation}
\begin{aligned}V_{\nu}(\boldsymbol{b}) & =\int d\Omega A_{\nu}(\hat{\boldsymbol{s}})\phi(\nu)I_{\nu}(\hat{\boldsymbol{s}})\exp\left[-2\pi i\frac{\nu}{c}\boldsymbol{b}\cdot\hat{\boldsymbol{s}}\right],\\
 & \approx\frac{2k_{B}}{\lambda^{2}}\int d\Omega A_{\nu}(\hat{\boldsymbol{s}})\phi(\nu)T(\hat{\boldsymbol{s}})\exp\left[-2\pi i\frac{\nu}{c}\boldsymbol{b}\cdot\hat{\boldsymbol{s}}\right],
\end{aligned}
\label{eq:Vis1}
\end{equation}
Here $\lambda$ is a mean wavelength, $\boldsymbol{b}$ is the baseline
length, $\hat{\boldsymbol{s}}$ and $\Omega$ are a direction in the
sky and its corresponding solid angle. $A_{\nu}$ is the (frequency
dependent) beam, and $I$ is the specific intensity, which has been
related to $T$, the brightness temperature in the Rayleigh-Jeans
limit. $k_B$ is Boltzmann's constant. Note that $A$ here is the baseline beam, i.e. product of two antenna voltage beams, examples of which are shown in Fig. \ref{fig:Beam}. $\phi(\nu)$ is the frequency bandpass profile. In practice,
power spectrum measurements are typically taken from a few ten MHz centered around the corresponding redshift of interest (e.g. 150 MHz for z=9.5). We take a moment to note that in Eq. \ref{eq:Vis1}, and our subsequent discussion, no flat-sky approximation has been made. The angular integral is performed over the dome $d\Omega$.

@@

Eq. \eqref{eq:Vb1} expresses the delay-transformed visibility as
an integral over observation coordinates $\hat{\boldsymbol{s}}$ and $\nu$. Ultimately,
we would like to relate the data, collected with coordinates $\hat{\boldsymbol{s}}$
and $\nu$, to the power spectrum, written with cosmological coordinates
$\boldsymbol{r}$ and $\boldsymbol{k}$. We start by noticing that
\[
\begin{aligned}r & =\frac{c}{H_{0}}\int_{0}^{z}\frac{dz'}{E(z')},\\
 & \approx\frac{c}{H_{0}}\int_{0}^{z_{0}}\frac{dz'}{E(z')}-\frac{c(1+z)^{2}}{\nu_{21}H_{0}E(z)}\left(\nu-\nu_{0}\right),\\
 & \equiv D_{c}-Y\Delta\nu,
\end{aligned}
\]
where $\nu_{21}=1420$MHz is the 21cm transition rest frequency, $\nu_{0}$
a reference central frequency with corresponding redshift $z_{0}$,
and 
\[
E(z)=\sqrt{\Omega_{m}(1+z)^{3}+\Omega_{\Lambda}}.
\]
Inverting for $\nu$:
\begin{equation}
\nu=\frac{D_{c}-r}{Y}+\nu_{21}.\label{eq:nur}
\end{equation}

We can thus rewrite the delayed transformed visibility as 
\small
\[
\begin{aligned}V(\boldsymbol{b},\tau) & =\frac{2k_{B}}{\lambda^{2}}\int_{H}\frac{d^{3}r}{X^{2}Y}A(\boldsymbol{r})\phi(r)T(\boldsymbol{r})\exp\left[-2\pi i\left(\frac{\boldsymbol{b}}{c}\cdot\hat{\boldsymbol{r}}+\tau\right)\nu_{r}\right]\end{aligned}.
\]
\normalsize
Here $(r_{x},r_{y},r_{z})=(X\hat{\boldsymbol{s}}_{x},X\hat{\boldsymbol{s}}_{y},Y\nu)$, and so $d\nu=-dr/Y$ and $d^{3}r=-X^{2}Yd\Omega d\nu$. 
$(Xk_{x},Xk_{y},Yk_{z})=\frac{2\pi}{c}(b_{x},b_{y},\tau)$ relate
the cosmological coordinates $r$ and $k$ to the measured coordinates.
We have written $\nu_{r}$ to remind us that $\nu$ and $r$ are related
by Eq. \eqref{eq:nur}. The beam reception pattern $A$ is dimensionless,
normalized to 1 at its peak (zenith), and we assume it to be the same
for all baselines.

>>>>

%\documentclass[preprint2,numberedappendix,tighten,twocolappendix]{aastex6}  % USE THIS TO MAKE BIB, THEN FORMAT USING EMULATEAPJ
\documentclass[twocolumn,apj,numberedappendix]{emulateapj}
\shorttitle{Adding Sensitivity to 21\,cm Inteferometric Probes of Reionization}
\shortauthors{Zhang\&Parsons}
\usepackage{float}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[figuresright]{rotating}
%\usepackage{rotating}
\usepackage{natbib}
%\usepackage{pdflscape}
%\usepackage{lscape}
\usepackage{ctable}
\citestyle{aa}
\renewcommand\[{\begin{equation}}
\renewcommand\]{\end{equation}}
\graphicspath{{../figures/}}

@@

\title{Adding Sensitivity to 21\,cm Inteferometric Probes of Reionization by Optimizing Choice of Baselines}

@@

\begin{abstract}
The observational effort to measure the cosmological 21\,cm power spectrum with radio interferometers requires high sensitivity. We introduce a straightforward and accurate method to extract sensitivity from visibility data. We develop our method in line with existing visibility-based power spectrum pipelines. Our method efficiently finds subsets of baselines to cross correlate, including partially redundant baselines, and quantifies the sensitivity contributions of given baselines. 
Using the 128-element configurations and beams of Precision Array for Probing the Epoch of Reionization (PAPER-128), as well as 4 planned versions of the Hydrogen Epoch of Reionization Array (HERA), we illustrate how our method applies to different arrays and predict the sensitivity improvements of including each baseline pairs. We show that inclusion of non-perfectly redundant baselines 
would account for $20\%$ to $60\%$ of the sensitivity of PAPER-128 and different configurations of HERA. 
\end{abstract}

@@

The epoch of reionization represents the last key
stage of our Universe's early evolution. Study of this event stands at
the intersection of cosmology and astrophysics. Understanding this
event not only serves as a scientific goal
of its own, but also as a gateway to crucial information
regarding fundamental physics of inflation, neutrino mass and phenomenology
of the first stars and galaxies (e.g. \citealt{LiuOpticalDepth, Liu2016b, Mao2008, DEw21cm, Bull2015, Oyama20131186} etc.).

Arguably the most promising observational probe of the epoch of reionization
comes from measurement of the ``spin-flip'' transition of neutral
hydrogen of characteristic wavelength 21\,cm \citep{Furlanetto2006181,PritchardLoeb}.
Other observational studies of reionization include for example Gunn-Peterson measurements of quasi-stellar objects \citep{Fan2006}. GP tests rely on Ly$\alpha$ absorption, and thus easily saturate with small degree of ionization and can only probe the end stage of reionization. The 21\,cm line can in comparison directly probe the
abundance and distribution of neutral hydrogen, and thus would potentially
shed light on all stages of reionization. Radio interferometric efforts
to measure the 21\,cm power spectrum has been a top priority in recent years of astronomy.
Current-generation instruments include the Precision Array for Probing
the Epoch of Reionization (PAPER) \citep{Ali2015,paper32}, Murchison
Widefield Array (MWA) \citep{Bowman2013, Tingay2013}, Low Frequency Array (LOFAR) \citep{LOFAR}. Next-generation instruments include the Hydrogen Epoch of Reionization
Array (HERA) (e.g. \citealt{HERA,HERAconfiguration,HERABEAM1,HERADISH2})  under construction, 
and the Square Kilometer Array Low (SKA-low) (e.g. \citealt{SKA1}), which is currently in planning.

These modern low-frequency radio interferometers aimed at measuring the 21\,cm signal bears designs dramatically different from traditional instruments built for imaging localized bright sources.  With the goal of detecting a cosmological signal 5 orders of magnitude fainter than foreground contaminations, current efforts are sensitivity starved. Maximum-redundancy arrays are thus designed to repeatedly measure the same Fourier signal to increase sensitivity. Ranging upwards from 100 elements, these arrays are typically of the drift-scan type to limit cost.

Analysis pipelines of such maximum-redundancy arrays typically fall into two categories. In the first, more traditional technique, image is formed in Fourier domain through rotation synthesis, then foreground is either removed or avoided to construct a power spectrum. The second, lately popular technique works directly with visibilities from baselines and cross multiply them to form the power spectrum. Due to its more natural focus on the instrumental output of visibilities, this technique is shown to fair better in dealing with systematics such as RFIs and instrumental cross-talks, and thus tend to provide good limits on the power spectrum. \cite{Ali2015} provides the newest upper limit to the power spectrum
measurements with the 64-element version of PAPER (henceforth referred to as PAPER-64). However, there is one caveat with existing visibility-based pipelines. While an imaging based power-spectrum pipeline naturally includes all redundancy information, visibility-based pipelines until now only cross multiply fully redundant baselines, in other words baselines of the same length and orientation.

Baselines of different length and orientation still contain partially redundant information. The use of partially redundant baselines is in principle not a fundamental limitation of visibility-based pipelines. In fact, most instrumental reports to date do include partial redundancy. Recently, \cite{wterm} proposed an algorithmic approach to extract power-spectrum from partially redundant baselines in tracking measurements, by carefully treating the $w$-term in the fringe pattern to obtain weights for cross-correlation. Our work parallels this effort by focusing on arrays of the drift-scan type that do not have tracking capabilities, such as PAPER, HERA and SKA.
The Earth's rotation causes the baselines in a drift-scan array to pick up different modes of the sky with time. Rotation-synthesis makes use of the rotation-induced $uv$ coverage map to form image. In a visibility based pipeline, we can extract the same rotation-induced redundancy. In this point of view, baselines that are slightly different in length and orientation
rotate into each other at a time delay. We can thus cross-multiply time-shifted visibilities, with the proper weighting, to form power spectrum.  Due to the large number of elements of modern arrays (upward of 100), the task of cross-multiplying every baseline against every other, scaling as number of array element to the fourth power, can be computationally formidable. Our contribution is thus twofold. First we introduce a formalism to include pairs of partially-redundant baselines in a visibility based power spectrum pipeline. Secondly we show how to use the formalism to automaticaly preselect baseline pairs and time offsets, making the problem computationally efficient. More precisely, our formalism allows one to simultaneously a) identify the baselines that give good
redundancy, b) find the time offset that corresponds to maximal redundancy for a given pair of baselines, and
c) quantify the sensitivity associated with cross multiplying
such a pair of partially redundant baselines, which in turns is used as weight to combine measurements in a power spectrum pipeline.

[TODO: this paragraph would potentially be promoted to a section] By using pairs of baselines, we also reduce additional mode mixing. As discussed in \cite{Hazelton2013}, the varying shapes of overlapping regions in $uv$ space of all contributing baselines lead to chromaticities in the response. By considering only pairs of baselines, we fix shapes of overlapping regions, and only let their sizes vary with frequency, just as in the case of fully redundant baselines.

The rest of this paper is organized as follows. In section 2 we introduce some terminology and notations to be used for the rest of the paper. In section
3 we introduce the formalism for weighting partially redundant baselines.
In section 4 we present numerical tests of
this technique as well as the expected sensitivity improvement
with this method for HERA and PAPER-128 and with section 5 we conclude.

\section{Notation and Terminology}

In order to eliminate confusion and ambiguity, we
introduce some terminology for the rest of this paper, some of which differs from what is commonly found in the literature.

First, we make the distinction of a \textit{``baseline''}, which refers to two specific antennas, and a \textit{``class of baselines''}, which refers to all baselines of the same length and orientation in a given array.

Baselines of the same class are traditionally called
``redundant baselines'', because they measure the same Fourier mode
in the sky.  We shall call baselines in the same class \textit{``equivalent baselines''}, and reserve the word \textit{``redundancy''} of two baselines in reference to a variable function of the relative time-offsets of their visibility time series. With this terminology, two equivalent
baselines are fully redundant with each other simultaneously at all
times. Non-equivalent baselines also have partial redundancy, which is maximized if their
respected time series are shifted with respect to one another by some delay. In other words, one baseline can be ``rotated into''
another if they are \textit{``near-equivalent''}.

\caption{The PAPER-128 layout. Each dot corresponds to the location of
an antenna. Top panel shows the antenna positions drawn to scale;
bottom panel show the antenna labels and distances, excluding the outrigger
antennas.
The numbering of the antennas the bottom panel are original labels
during instrument assembly and does not bear significant
meaning. In the bottom panel, the two baselines (5,77) and (40,85) are an example of equivalent pair, with separation denoted by \{1, 0\}, for the
antennas are separated by 1 unit east and 0 unit north. Similarly,
the baselines (5,69) and (5,85) are examples
of the classes \{1, 1\} and \{1,-1\}, respectively. Note \{1, 0\} and \{-1, 0\} for example
are the same baseline classes and should not be counted twice.}
\label{fig:AntPos}
\end{figure}

We shall use the 128-element PAPER array (henceforth as PAPER-128) to motivate our formalism and demonstrate our method, and extend our results to several HERA configurations in Sec \ref{sec:arrconf}. 
The PAPER array is located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern with antenna labels are show
in Fig. \ref{fig:AntPos}. The array consists of a 112-element grid and 16 "outriggers" used primarily to aid calibration. We see the antenna spacing in North-South
directions are comparatively close (4m), so that baselines such as
(0, 44) and (0, 7) are very close to equivalent. In the bottom panel, the two baselines (5, 77) and (40, 85) are an example of equivalent pair, with separation denoted by \{1, 0\}, for the
antennas are separated by 1 unit east and 0 unit north. Similarly,
the baselines (5, 69) and (5, 85) are examples
of \{1, 1\} and \{1, -1\}, respectively.
Note \{1, 0\} and \{-1, 0\} for example are the same classes and should
not be counted twice. Antennas in purely north-south baselines
are close enough to induce cross-talk, and hence are not suitable
for use. The PAPER-64 analysis of \cite{Ali2015} used three classes of baselines, the PAPER-128
equivalent of which are 
\{2, 0\}, \{2, 1\} and \{2, -1\}. There each of these classes
of baselines are cross multiplied to itself.

@@

Given a point source on the sky, each baseline maps the
source to a point in $uv$ plane. As the Earth rotates with respect to
the source, the point traces out tracks in the $uv$ plane. 
We show in Fig. \ref{fig:Tracks} $uv$ tracks of PAPER-128 over 4.8 hours, at 0.15GHz, for a source that passes through zenith.

\begin{figure}[H]
\includegraphics[width=\linewidth]{tracks128}
\caption{Tracks of PAPER-128 (grid only, excluding outriggers) for requency $\nu=0.15\text{GHz}$ and a hypothetical source that passes through zenith.
These tracks are traced out over 0.2 sidereal days, or roughly 4.8
hours. Color represents different baselines. 
As the Earth rotates, tracks are traced out counterclockwise. }
\label{fig:Tracks}
\end{figure}

@@

We take the visibility as commonly defined in the literature (e.g.
\citealt{first-paper}): 
\begin{equation}
\begin{aligned}V_{\nu}(\boldsymbol{b}) & =\int d\Omega A_{\nu}(\hat{\boldsymbol{s}})\phi(\nu)I_{\nu}(\hat{\boldsymbol{s}})\exp\left[-2\pi i\frac{\nu}{c}\boldsymbol{b}\cdot\hat{\boldsymbol{s}}\right],\\
 & \approx\frac{2k_{B}}{\lambda^{2}}\int d\Omega A_{\nu}(\hat{\boldsymbol{s}})\phi(\nu)T(\hat{\boldsymbol{s}})\exp\left[-2\pi i\frac{\nu}{c}\boldsymbol{b}\cdot\hat{\boldsymbol{s}}\right],
\end{aligned}
\label{eq:Vis1}
\end{equation}
Here $\lambda$ is a mean wavelength, $\boldsymbol{b}$ is the baseline
length, $\hat{\boldsymbol{s}}$ and $\Omega$ are a direction in the
sky and its corresponding solid angle. $A_{\nu}$ is the (frequency
dependent) beam, and $I$ is the specific intensity, which has been
related to $T$, the brightness temperature in the Rayleigh-Jeans
limit. $k_B$ is Boltzmann's constant. Note that $A$ here is the baseline beam, i.e. product of two antenna voltage beams, examples of which are shown in Fig. \ref{fig:Beam}. $\phi(\nu)$ is the frequency bandpass profile. In practice,
power spectrum measurements are typically taken from a few ten MHz centered around the corresponding redshift of interest (e.g. 150 MHz for z=9.5). We take a moment to note that in Eq. \ref{eq:Vis1}, no flat-sky approximation has yet been made. The angular integral is performed over the dome $d\Omega$.

@@

Eq. \eqref{eq:Vb1} expresses the delay-transformed visibility as
an integral over observation coordinates $\hat{\boldsymbol{s}}$ and $\nu$. Ultimately,
we would like to relate the data, collected with coordinates $\hat{\boldsymbol{s}}$
and $\nu$, to the power spectrum, written with cosmological coordinates
$\boldsymbol{r}$ and $\boldsymbol{k}$. We start by noticing that
\[
\begin{aligned}r & =\frac{c}{H_{0}}\int_{0}^{z}\frac{dz'}{E(z')},\\
 & \approx\frac{c}{H_{0}}\int_{0}^{z_{0}}\frac{dz'}{E(z')}-\frac{c(1+z)^{2}}{\nu_{21}H_{0}E(z)}\left(\nu-\nu_{0}\right),\\
 & \equiv D_{c}-Y\Delta\nu,
\end{aligned}
\]
where $\nu_{21}=1420$MHz is the 21\,cm transition rest frequency, $\nu_{0}$
a reference central frequency with corresponding redshift $z_{0}$,
and 
\[
E(z)=\sqrt{\Omega_{m}(1+z)^{3}+\Omega_{\Lambda}}.
\]
Inverting for $\nu$:
\begin{equation}
\nu=\frac{D_{c}-r}{Y}+\nu_{21}.\label{eq:nur}
\end{equation}

Further we shall invoke the flat-sky limit
\begin{equation}
(r_{x},r_{y})=(X\hat{\boldsymbol{s}}_{x},X\hat{\boldsymbol{s}}_{y}), 
\end{equation}
where
\begin{equation}
X = \frac{c}{H_0}\int_{0}^{z}\frac{dz'}{E(z')}
\end{equation}
is a cosmological scalar. In the same limit $(Xk_{x},Xk_{y},Yk_{z})=\frac{2\pi}{c}(b_{x},b_{y},\tau)$ relates
the cosmological $k$ to the observation coordinates. For a more rigorous treatment without the flat sky limit, see Appendix [ADRIAN].

We can rewrite the delayed transformed visibility as 
\small
\[
\begin{aligned}V(\boldsymbol{b},\tau) & =\frac{2k_{B}}{\lambda^{2}}\int_{H}\frac{d^{3}r}{X^{2}Y}A(\boldsymbol{r})\phi(r)T(\boldsymbol{r})\exp\left[-2\pi i\left(\frac{\boldsymbol{b}}{c}\cdot\hat{\boldsymbol{r}}+\tau\right)\nu_{r}\right]\end{aligned},
\]
\normalsize
 where $d\nu=-dr/Y$ and $d^{3}r=-X^{2}Yd\Omega d\nu$.

We have written $\nu_{r}$ to remind us that $\nu$ and $r$ are related
by Eq. \eqref{eq:nur}. The beam reception pattern $A$ is dimensionless,
normalized to 1 at its peak (zenith), and we assume it to be the same
for all baselines.

With a time offset, the beam pattern has moved relative to the sky.
Here we choose to fix the sky, and denote the rotated coordinates
of the beam pattern with the 3 dimensional rotation operator $\Gamma$. With implicit bounds
of integrals from $-\infty$ to $\infty$, we have:
\begin{widetext}
\begin{equation}
\begin{aligned} & \langle V^{*}(\boldsymbol{b},\tau)V(\boldsymbol{b'},\tau')\rangle\\
 & =\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}\int d^{3}rd^{3}r'\left(\langle T^{*}(\boldsymbol{r})T(\boldsymbol{r'})\rangle\right)A^{*}(\boldsymbol{r})A(\Gamma \boldsymbol{r'})\Phi_{b,\tau}(\boldsymbol{r},\Gamma \boldsymbol{r'}),\\
 & =\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}\int d^{3}rd^{3}r'\left(\int\frac{d^{3}\kappa}{(2\pi)^{3}}\frac{d^{3}\kappa'}{(2\pi)^{3}}\langle T^{*}(\boldsymbol{\kappa})T(\boldsymbol{\kappa'})\rangle e^{-i(\boldsymbol{\kappa}\cdot \boldsymbol{r}-\boldsymbol{\kappa'}\cdot\boldsymbol{r'})}\right)A^{*}(\boldsymbol{r})A(\Gamma \boldsymbol{r'})\Phi_{b,\tau}(\boldsymbol{r},\Gamma \boldsymbol{r'}),\\
 & =\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}\int d^{3}rd^{3}r'\left(\int\frac{d^{3}\kappa}{(2\pi)^{3}}P(\kappa)e^{-i\boldsymbol{\kappa}\cdot(\boldsymbol{r}-\boldsymbol{r'})}\right)A^{*}(\boldsymbol{r})A(\Gamma \boldsymbol{r'})\Phi_{b,\tau}(\boldsymbol{r},\Gamma \boldsymbol{r'}),\\
 & \approx\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}P(k_{b,\tau})\int d^{3}rd^{3}r'\delta_{D}^{(3)}(\boldsymbol{r}-\boldsymbol{r'})A^{*}(\boldsymbol{r})A(\Gamma \boldsymbol{r'})\Phi_{b,\tau}(\boldsymbol{r},\Gamma \boldsymbol{r'}),\\
 & =\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}P(k_{b,\tau})\int d^{3}r|A^{*}(\boldsymbol{r})A(\Gamma \boldsymbol{r})||\phi(\nu_{r})|^{2}\exp\left[-i2\pi\nu_{r}\left(\hat{\boldsymbol{r}}\cdot\frac{\boldsymbol{b}}{c}-\Gamma \hat{\boldsymbol{r}}\cdot\frac{\boldsymbol{b'}}{c}\right)\right],\\
 & =\left(\frac{2k_{B}}{\lambda^{2}}\right)^{2}P(k_{b,\tau})\int\frac{d\Omega d\nu}{X^{2}Y}|A^{*}(\hat{\boldsymbol{s}},\nu)A(\Gamma \hat{\boldsymbol{s}},\nu)||\phi(\nu)|^{2}\exp\left[-i2\pi\nu\left(\hat{\boldsymbol{s}}\cdot\frac{\boldsymbol{b}}{c}-\Gamma\hat{\boldsymbol{s}}\cdot\frac{\boldsymbol{b'}}{c}\right)\right],
\end{aligned}
\label{eq:main}
\end{equation}

where in transition from cosmological coordinates back to observing coordinates we have written $\hat{\boldsymbol{r}}\equiv\hat{\boldsymbol{s}}$, and 
\begin{equation}
\Phi_{b,\tau}(\boldsymbol{r},\Gamma \boldsymbol{r'})=|{\phi^{*}}(\ensuremath{\nu_{r}})\phi(\nu_{r'})|\exp\left[-i\frac{2\pi}{c}\left(\boldsymbol{b}\cdot\nu_{r}\hat{\boldsymbol{r}}-\boldsymbol{b'}\cdot\nu_{r'}\Gamma\hat{r'}\right)\right]\exp\left[-i2\pi\tau\left(\nu_{r}-\nu_{r'}\right)\right].
\end{equation}
\end{widetext}
The second to third line of Eq.(\ref{eq:main}) follows from assumption of translational invariance, and the third to fourth line follows from the assumption that
the 3D power spectrum varies negligibly over the $k$-space of interest
so that $\hat{P}_{21}(k+k_{2})\approx\hat{P}_{21}(k)$. Since $\Gamma$
is a sky rotation, it does not affect $\nu$, hence we have taken $\nu_{r}$
outside the parenthesis. Notice that the phase factor $\exp\left[-i2\pi\tau\left(\nu-\nu'\right)\right]$
drops out in the end. This means that correlation is peaked at the same time-offset for all delay channels.
